Detection of eco-evolutionary dynamics in metacommunities using Joint Species

### 2024-11-30 manuscript/manuscript_v13.Rmd
Updated mscript, all appendices, all figures.

---
title: "Detection of eco-evolutionary dynamics in metacommunities using Joint Species
  Distribution Models"
author:
- "Jelena H. Pantel^[Laboratoire Chrono-environnement,UMR 6249 CNRS-UFC, 16 Route
  de Gray, 25030 Besançon cedex, France, jelena.pantel@univ-fcomte.fr]"
- "Ruben J. Hermann^[University of Duisburg-Essen, Universitätsstraße 5, 45141 Essen,
  Germany, ruben.hermann@uni-due.de]"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  word_document:
    reference_docx: docx_template.docx
  bookdown::html_document2: default
  bookdown::pdf_document2:
    toc: no
    fig_caption: yes
    latex_engine: xelatex
    extra_dependencies: float
    number_sections: yes
  pdf_document:
    toc: no
header-includes:
- \usepackage{fancyhdr}
- \pagestyle{fancy}
- \fancyhead[L]{Author Pantel and Hermann}
- \fancyhead[R]{Detection of ecoevo dynamics}
- \usepackage{lineno}
- \linenumbers
linestretch: 2
link-citations: yes
linkcolor: blue
csl: "the-american-naturalist.csl"
bibliography: refs.bib
---

@@

XXXXX

@@

We evaluated whether linear regression models can appropriately capture the dynamics that drive species abundances in relatively simple populations and communities. We fit data generated from simulation models with increasing complexity (\@ref(tab:tab1)) to Bayesian linear regression models. These simulation models were (A) a single-species model of discrete-time logistic population growth, (B) expanded to include a randomly fluctuating environmental property $E$ that impacts the species' population growth rate, (C) a two-species model of logistic growth and competition that was then (D) expanded to consider a randomly fluctuating environmental property, (E) and further expanded to consider evolution in a trait $x$ that is selected by the environmental property, and finally (F) the model of two-species growth, competition, environmental change, and evolution in a spatially structured (10-patch) habitat.

@@

To determine the relative importance of evolution for community composition, we fitted two HMSC models to the data: An 'evolution model' in which $\Delta \hat{x}$ of each species and environmental conditions were used as explanatory variables, and a 'no-evolution model', in which only the environmental conditions were included. Species density was log-transformed species before applying it to the model, therefore we could use normal distribution in the model. We also included replicates (*n* = 2) and time as random effects. For each model we ran four MCMC chains to generate 1000 samples of the posterior distribution for all model parameters. To evaluate the model-estimated role of evolution for predicted population size, we used the posterior distributions of the estimated effect size of $|\Delta x|$ multiplied by the observed $\Delta \hat{x}$ of each species,added to the model-estimated intercept.

@@

When a second species was added (models *C* and *D*), the regression coefficients performed well for predicting density values at all time points, but the direction and magnitude of the regression coefficients did not reflect the intra- and interspecific competition strengths included in the mechanistic model (Appendix S1.3). We instead used a latent variable modelling approach (implemented in HMSC) to consider these species interactions as part of the temporal random effects included in the model. This approached produced effect size estimates that both predicted the population dynamics well (Figure 1a \@ref(fig:fig1)) and captured the conditional effects of environment for species densities (Figure 1b \@ref(fig:fig1)). The variation partition also successfully captured the relative importance of the environmental predictor and the temporal random effects (which in this simple case were driven by intra- and interspecific species interactions; Figure 1c \@ref(fig:fig1); Appendix S1.3).

This approach produced estimates of main effects of environment (Figure 1a,b)

However, as the processes included in the simulation models become slightly more complex (competition with a second species, spatial structure; models *C*, *D*, *E*, *F*), regression coefficients were useful for prediction but not for recreating the underlying processes driving changes in population size (Appendix S1).

To consider species interactions and environmental fluctuations (model *D*; Figure 1a \@ref(fig:fig1)), we used the latent variable modelling approach implemented in HMSC. The linear model successfully estimated coefficients for fixed environmental effects that captured the impacts of environment for population size in both species (Figure 1b) and partitioned the relative importance of the environment and species interactions for species abundances (Figure 1c). The model coefficients could also predict the population size values over time (Figure 1a), though the quality of this prediction varied depending on the level of environmental noise (Appendix S1).

```{r fig1, out.width="90%", include=TRUE, fig.align="center", fig.cap=c("Simulated and model-estimated values for population dynamics in two-species systems. Panels a, d, and g show log-population size (y-axis) over time (x-axis) for species $i$ and $j$ when they grow and compete for resources in sites with random environmental fluctuations over time. Species either have fixed (panel a) or evolving trait values (panels d, g) and are present in either 1 (panels a, d; simulation models D and E) or multiple sites (panel g; model F). Graphs in panels b, e, and h show predicted log-population size ($ln(N)$) for species 1 and 2 across gradients of the environment ($E + E^2$) and the change in trait values of species 1 ($|\\Delta x_1|$). Graphs in panels c, f, and i show the results of a variation partition after fitting data to the HMSC statistical model. The variation partition plot shows the proportion of variation explained by environmental, trait evolution ($|\\Delta x_1|$ and $|\\Delta x_2|$), time random, and spatial random predictors for $ln(N)$ for species 1 and 2. Error bars and shading represent the 95% highest probability density interval for posterior predictions."), echo=FALSE}
knitr::include_graphics("../output/fig1.pdf")
```

The HMSC model successfully modelled changes in species abundances when trait evolution was introduced (model *E*; Figure 1d \@ref(fig:fig1)). Trait evolution played a strong role for population size, especially the effect of $|\Delta x_1|$ for $ln(N_1)$ and $ln(N_2)$ (gradient plots in Figure 1e). The model estimated the relative importance of the environmental driver, trait evolution in both species, and species interactions (Figure 1f), and predicted the observed species abundances well (Figure 1d). Spatial structure was also reproduced in the coefficients of the linear HMSC model by using spatial random effects to capture covariance in species abundances as a function of spatial distance between sites in the 10-patch simulation model (model *F*; Figure 1g,h,i \@ref(fig:fig1)). The variation partition successfully reflected separation of environmental drivers from trait evolution, species interactions, and spatial random effects (Figure 1i), and we show in Appendix S1 that the relative importance of trait evolution increases when we use a simulation model with less spatial and environmental variation and weak species interactions.

@@

The simulation model followed population size and trait value dynamics for species in the landscape across 200 time steps (Figure 2a \@ref(fig:fig2)). Resulting biodiversity dynamics depended on the dispersal level and heritability values (Figure 2b \@ref(fig:fig2)). For dispersal, results mirrored some expectations from existing metacommunity models, i.e. that regional $\gamma$ diversity decreases with increasing dispersal level ([@Mouquet2003]; Figure 2b \@ref(fig:fig2)). The inclusion of trait heritability had a strong effect on community diversity, reflecting that local adaptation can rescue species and contribute to richness at the local and metacommunity scale. Regional species diversity ($\gamma$) increased with increasing $h^2$ levels (Figure 2b \@ref(fig:fig2)).

Analysis using the HMSC model for community time series across $d$ and $h^2$ levels indicated that the importance of evolution for community composition ($V_{|\Delta x|}$) decreased with increasing dispersal and with increasing heritability (Figure 2c,d,e \@ref(fig:fig2)). The importance of spatial random effects increased with dispersal rate, and overall the importance of evolution decreased with increasing $h^2$.

```{r fig2, out.width="90%", include=TRUE, fig.align="center", fig.cap=c("Importance of trait evolution for change in community composition in metacommunities. (a) We developed a simulation model for population growth, competition, and trait evolution in 15 species that inhabit a patchy landscape. The landscape has 50 sites (colored points in the XY space), with distinct local environments that select for diverse patterns of species diversity (size of points) and average trait values (color of points; results are shown at time $t=0, 10, 200$ for a simulation with $d=10^-3$ and $h^2 = 0.1$). (b) Regional metacommunity $\\Gamma$ diversity varied across levels of dispersal and heritability ($h^2$). (c) Variation partition plots after fitting metacommunity time series of species abundances $ln(N_{ikt})$ to HMSC models with environment, trait evolution, and random effects (space and time) as predictors. The top panel is for $h^2 = 0.1$, the middle panel is for $h^2 = 0.3$, and the bottom panel is for $h^2 = 0.9$."), echo=FALSE}
knitr::include_graphics("../output/fig2.pdf")
```

@@

The empirical data [@Jewell2023] showed the community dynamics of four plant species under different environmental conditions, additionally the front size of each species was measured to determine its trait change over time. We calculated the mean variance partitioning (VP) over all species, to evaluate the impact of evolutionary (change of the front size trait) and ecological (light and nutrient treatments) effects on changes on species density. Almost one fifth (18.6%) of the variance in species density could be explained by the variance in $\Delta \hat{x}$ (\@ref(fig:fig5)a). This is due to the reduction of the VP of the ecological effects by a total of 14.4% (from 37% to 26.6% for the amount of shading and from 31% to 27.1% for the nutrient concentrations) and the reduction of 4.2% of the random effects (from 18.7% to 15% for time and from 13.3% to 12.8% for replicates) (\@ref(fig:fig5)). While $\Delta \hat{x}$ of the species Lm, Sp and Wc showed a similarly low impact, on average 3.6%, $\Delta \hat{x}$ of species Rn showed a high impact on variance of species density, 7.8%, which can be explained by its high $\Delta \hat{x}$ changes over time (see Appendix S05). The high density intervals (HDI) of 95% estimated beta effects showed a negative correlation of Light shading with species density, while nutrient concentration correlated positively with density (\@ref(fig:fig5)c). Changes in $\Delta \hat{x}$ impacted densities differently. At one hand changes of species Rn frond size correlated negatively with is density, on the other hand changes in the frond size of species Wc showed a positive correlation with the other species Sp and Lm (\@ref(fig:fig5)c). The impact of evolution (of all species summed up), impacted species density over time, yet did not show a clear direction over the course of the experiment (\@ref(fig:fig5)d).

```{r fig3, out.width="90%", include=TRUE, fig.align="center", fig.cap=c("The estimated effects of species trait changes and environmental treatments. The variance partitioning of all fixed effects of the HMSC model (a) including trait changes with envrionmental treatments and (b) only including the environmental treamtents. The (b) high density interval of 95% showed a clear pattern for the different environmental treatments, yet differing results for trait changes and their impact on species density. This can be seen in (d) the shifting predictions of density over time due to the total trait changes of all species."), echo=FALSE}
knitr::include_graphics("../output/fig5.pdf")
```

@@

*Main findings*
The analysis of community composition often requires statistical models that can consider the combined influences of environmental drivers and spatial connectivity. A more difficult, yet critical, element to incorporate is species interactions. Although the causal role of species interactions is difficult to interpret from statistical models (but see [@Dubart2022] [@Luo2022]) in systems with few species), joint species distribution models (JSDMs) have implemented spatial and temporal random effects that can effectively capture these species interactions using a latent variable approach (but note the resulting species covariance matrices will also reflect other processes besides species interactions; [@Poggiato2021]). Finally, one element that has not been fully implemented into joint species distribution models is dynamic trait evolution (see [@Garzon2019] for a model that does consider reaction norms from in-situ reciprocal transplant experiments). We used an existing JSDM and developed simulation models of evolving populations in increasing complex scenarios to determine whether the signal of trait evolution could be estimated from data and compared to other drivers of community structure. We also applied this statistical model (HMSC, hierarchical models of species distributions; [@Ovaskainen2017HMSC] [@Tikhonov2020]) to an existing empirical dataset (from the experimental study of community eco-evolutionary dynamics in aquatic plants of [@Jewell2023]) to show how HMSC can be used to model the changes in species composition over time as a function of environmental drivers, trait change, and temporal random effects. Overall, there is no barrier to including dynamic trait change as a fixed effect in a JSDM model, and the resulting effect size coefficient estimates can be used to (i) estimate the relative importance of trait evolution for community composition and (ii) predict the impacts of future trait change in the system.

*Discussion regarding simple case studies*
At the same time, fixed effects for the impacts of trait evolution on population size should be interpreted with a great deal of caution. For example, our simulations reveal instances where trait evolution may occur at the same time as large changes in population size, and this co-occurrence isn't always causal (e.g. when selection is weak but trait heritability is high, and species are introduced at low initial population size). In this instance, the regression coefficient from the model may be negative and strong. By using simulations of trait evolution and population growth in simple systems (1- and 2-species systems, at single and simplified multi-patch scales; \@ref(fig:fig1); Appendix S1), we can better interpret effect size estimates. However, these precautions are not unique to including trait evolution as a predictor of community composition. For example, previous research has shown that the effect size coefficients from linear and AR models using $N_{i,t-1}$ to predict $N_{i,t}$ or $N_{j,t}$ are not equivalent to the intra- and interspecific interaction coefficients that determine species density dependence and should not automatically be used to interpret the direction and magnitude of species interactions ([@Kloppers2013] [@Certain2018] [@Muhlbauer2020] [@Olivenca2021]). Our findings for using HMSC to estimate the impact of trait evolution ($|\Delta x_i|$) for species abundances are similar - (i) the relative importance of trait evolution estimated by a variation partition accurately reflects the relative strength of simulated species interactions vs. magnitude of trait change and (ii) the coefficients greatly increase the predictive ability of the statistical model for simulated data (Appendix S1). HMSC is therefore a useful tool for understanding the role that evolution can play in community dynamics.

*Discussion regarding metacom simulations*
This was also observed when using HMSC to evaluate the importance of evolution for metacommunity structure, using simulated data from a model of population growth, competition, dispersal, and trait evolution in a spatially structured habitat (Figure 2a \@ref(fig:fig1); Appendix S4). First of all, the simulations themselves revealed an intriguing role of adaptive evolution for rescuing individual species from extinction (i.e. [@Gomulkiewicz1995]) and as a consequence greatly increasing biodiversity at the local and the landscape scale (i.e. community rescue, [@Low2015] [@Bell2019]; Figure 2b \@ref(fig:fig2)). The recovery of species via local adaptation had consequent effects for coexistence, as species that were initially mal-adapted could recover in population size and exert strong competitive effects on other species. Because HMSC can accurately estimate the relative importance of trait evolution for community composition (Appendix S4), we were thus able to test hypotheses about the drivers of increased or decreased impacts of trait evolution for community structure. Our simulation models fit to HMSC confirmed our hypotheses that the importance of evolution would decrease with increasing landscape connectivity (modeled as species dispersal rate $d$) and the overall speed of adaptive evolution (modeled as heritability $h^2$). Previous theoretical models have been shown that these two properties can have a strong influence on resulting community composition ([@Urban2008] [@Loeuille2008] [@Urban2009] [@DeMeester2016]), but our demonstration that the HMSC statistical model can accurately estimate $R^2$ values for the relative importance of trait evolution can allow hypotheses from theoretical models to be tested using empirical data and compared across different study systems.

*Discussion regarding empirical data*
By implementing the trait changes of the four different species (Lm, Rn, Sp and Wc) in the HMSC model we were able to infer the relative impact of the trait changes on the community dynamics. This application shows is advantages over previous methods,which estimate the impact of trait changes (thus evolution), as it is not depended on the data structure and simplicity [@Hairston2005] [@Ellner2011], can be implemented on multiple species[@Lasky2020] and does not require additional common garden experiments [@Garzon2019]. Therefore this allows for a straight forward implementation of HMSC on a variety of data structures as lone as a clear response variable (e.g. species density) and ecological (e.g. environmental conditions) as well as evolutionary (e.g. heritable trait change) are measured. An additional advantage of HMSC, and JSDMs in general, is the ability to use the empirical data to predict a species response to changes in the environmental variables, even beyond the measured values [@Tikhonov2020] [@Ovaskainen2015]. With the implementation this approach can be combined with the fixed effects to answer further questions, such as: How does trait change impact the species density under changing environmental conditions? And how does the impact of trait changes depend on specific environmental conditions?

At the same time a correct interpretation of such predictions as well es the estimated posterior beta distribution require a thorough understand of the experimental system. For example we often observed a negative correlation between trait change and species density in our simulations and also in species Rn of the empirical data, yet they derived from different mechanisms. In our simulations trait changes were greater when a species trait was further away from the optimal trait value, this greater distance resulted also in lower fitness and thus lower population size, while a species with higher fitness and thus higher population size had smaller trait changes. Therefore, HMSC correlates a high trait changes with a low population size and thus estimates a negative beta effect. Such effects can be expected in empirical data of i.e. evolutionary rescue experiments, where a species has to adapt to a deleterious environment before going extinct [@Gomulkiewicz1995][@Bell2009]. In the empirical data this negative correlation was the result of the species Rn investing into a larger frond size (Fig. SX), this likely lead to a lower investment into reproduction, while still maintaining a similar total frond area on the water surface to capture the reduced light.

>>>>

---
title: "Detection of eco-evolutionary dynamics in metacommunities using Joint Species
  Distribution Models"
author:
- "Jelena H. Pantel^[Laboratoire Chrono-environnement,UMR 6249 CNRS-UFC, 16 Route
  de Gray, 25030 Besançon cedex, France, jelena.pantel@univ-fcomte.fr]"
- "Ruben J. Hermann^[University of Duisburg-Essen, Universitätsstraße 5, 45141 Essen,
  Germany, ruben.hermann@uni-due.de]"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document:
    toc: no
  bookdown::html_document2: default
  bookdown::pdf_document2:
    toc: no
    fig_caption: yes
    latex_engine: xelatex
    extra_dependencies: float
    number_sections: yes
    keep_tex: true
  word_document:
    reference_docx: docx_template.docx
  html_document:
    toc: no
    df_print: paged
header-includes:
- \usepackage{fancyhdr}
- \pagestyle{fancy}
- \fancyhead[L]{Author Pantel and Hermann}
- \fancyhead[R]{Detection of ecoevo dynamics}
- \usepackage{lineno}
- \linenumbers
linestretch: 2
link-citations: yes
linkcolor: blue
csl: "the-american-naturalist.csl"
bibliography: refs.bib
---

@@

Biodiversity at the metacommunity scale is typically influenced by a number of environmental, spatial, biotic, and stochastic factors. At the same time, these factors impact the evolution of individual species, as sites present different local selection pressures and connectivity can impact gene flow and genetic drift. Identifying the relative impacts of environmental, spatial, biotic, and other drivers on community composition across spatial and temporal scales has been greatly facilitated by joint species distribution models, but these models have yet to consider the impact of microevolution on community composition. We used Heirarchical Models of Species Communities (HMSC) to analyze simulated data of an populations and communities, including in a large evolving metacommunity model, to establish whether HMSC can sufficiently quantify the contribution of phenotypic evolution for metacommunity composition. The models successfully partitioned variance contributed by environmental, spatial, and evolving phenotypic drivers, and also estimated site- and year-specific covariance. We also applied the HMSC with trait evolution model to an existing dataset studying trait change and community dynamics in experimental aquatic plant system. The study of eco-evolutionary dynamics may require data that reflects numerous complex, interacting processes and it is necessary to have flexible, generalized statistical models to analyze this data. JSDM models such as HMSC present one promising path for analysis of eco-evolutionary dynamics in multi-species communities.

@@

We evaluated whether linear regression models can appropriately capture the dynamics that drive species abundances in relatively simple populations and communities. We fit data generated from simulation models with increasing complexity (Table \@ref(tab:tab1)). to Bayesian linear regression models. These simulation models were (A) a single-species model of discrete-time logistic population growth, (B) expanded to include a randomly fluctuating environmental property $E$ that impacts the species' population growth rate, (C) a two-species model of logistic growth and competition that was then (D) expanded to consider a randomly fluctuating environmental property, (E) and further expanded to consider evolution in a trait $x$ that is selected by the environmental property, and finally (F) the model of two-species growth, competition, environmental change, and evolution in a spatially structured (10-patch) habitat.

@@

To determine the relative importance of evolution for community composition, we fitted two HMSC models to the data: An 'evolution model' in which $|\Delta x|$ of each species and environmental conditions were used as explanatory variables, and a 'no-evolution model', in which only the environmental conditions were included. Species density was log-transformed species before applying it to the model, therefore we could use normal distribution in the model. We also included replicates (*n* = 2) and time as random effects. For each model we ran four MCMC chains to generate 1000 samples of the posterior distribution for all model parameters. To evaluate the model-estimated role of evolution for predicted population size, we used the posterior distributions of the estimated effect size of $|\Delta x|$ multiplied by the observed $\Delta x$ of each species, added to the model-estimated intercept.

@@

When a second species was added (models *C* and *D*), the regression coefficients performed well for predicting density values at all time points, but the direction and magnitude of the regression coefficients did not reflect the intra- and interspecific competition strengths included in the mechanistic model (Appendix S1.3). We instead used a latent variable modelling approach (implemented in HMSC) to consider these species interactions as part of the temporal random effects included in the model. This approached produced effect size estimates that both predicted the population dynamics well (Figure \@ref(fig:fig1) a) and captured the conditional effects of environment for species densities (Figure \@ref(fig:fig1) b). The variation partition also successfully captured the relative importance of the environmental predictor and the temporal random effects for model *D* (which in this simple case were driven by intra- and interspecific species interactions; Figure \@ref(fig:fig1) c; Appendix S1.3, S1.4).

Trait evolution was included as a predictor when it was introduced in model *E*. The resulting linear regression coefficients could predict species dynamics (Figure \@ref(fig:fig1) d) and reflected the impact and magnitude of environmental and trait evolution predictors (Figure \@ref(fig:fig1) e). The variation partition also showed that trait evolution in species 1 ($|\Delta x_1|$) was a strong driver of variation in abundance (blue portion, Figure \@ref(fig:fig1) f), primarily for species 1 but also for species 2 as well (Appendix S1.5; Appendix S2).

After adding spatial complexity to the 2-species simulation, we included a spatial random effect to capture species covariances as a function of distance between sites (model *F*). By properly considering species interactions as a temporal random effect and including spatial random effects, the HMSC model could estimate regression coefficients for the impacts of the environment $E_t$ and trait evolution $|\Delta x_1|$ and $|\Delta x_2|$ for both species that fit well to the simulated data points (Figure \@ref(fig:fig1 g); Appendix S1.6) and capture conditional effects of environment and trait evolution for both species (Figure \@ref(fig:fig1) h). Spatial random effects explained a large proportion of variance in abundance for both species, but trait evolution remained a strong driver for species 1.

```{r fig1, out.width="110%", include=TRUE, fig.align="center", fig.cap=c("Simulation models for 2-species ecological and eco-evolutionary dynamics. Panels a, d, and g show log-population size (y-axis) over time (x-axis) for species $1$ and $2$ when they grow and compete for resources with random environmental fluctuations over time. Species either have fixed (panel a) or evolving trait values (panels d, g) and are present in either 1 (panels a, d; simulation models D and E) or multiple sites (panel g; model F). Graphs in panels b, e, and h show predicted log-population size ($ln(N)$) for species 1 and 2 across gradients of the environment ($E + E^2$) and the change in trait values of species 1 ($|\\Delta x_1|$). Graphs in panels c, f, and i show the results of a variation partition after fitting data to the HMSC statistical model. The variation partition plot shows the proportion of variation explained by environment, trait evolution ($|\\Delta x_1|$ and $|\\Delta x_2|$), and temporal and spatial random effects for $ln(N)$ for species 1 and 2. Error bars and shading in panels a, d, and g represent the 95% highest probability density interval for posterior predictions."), echo=FALSE}
knitr::include_graphics("../output/fig1_mod.pdf",error=FALSE)
```

@@

The simulation model followed population size and trait value dynamics for species in the landscape across 200 time steps (Figure \@ref(fig:fig2) a). Resulting biodiversity dynamics depended on the dispersal level and heritability values (Figure \@ref(fig:fig2) b). For dispersal, results mirrored some expectations from existing metacommunity models, i.e. that regional $\gamma$ diversity decreases with increasing dispersal level $d$ ([@Mouquet2003]; Figure \@ref(fig:fig2) b). The inclusion of trait heritability had a strong effect on community diversity, reflecting that local adaptation can rescue species and contribute to richness at the local and metacommunity scale. Regional species diversity ($\gamma$) increased with increasing $h^2$ levels (Figure \@ref(fig:fig2) b).

Analysis using the HMSC model for community time series across $d$ and $h^2$ levels indicated that the importance of evolution for community composition ($V_{|\Delta x|}$) decreased with increasing dispersal and with increasing heritability (Figure \@ref(fig:fig2) c). The importance of spatial random effects increased with dispersal rate, and overall the importance of evolution decreased with increasing $h^2$. These effects can be better understood by our supplemental analyses of 1-species, 1-site simulation models of evolutionary rescue (Appendix S3).

```{r fig2, out.width="110%", include=TRUE, fig.align="center", fig.cap=c("Importance of trait evolution for change in community composition in metacommunities. (a) We developed a simulation model for population growth, competition, and trait evolution in 15 species that inhabit a patchy landscape. The landscape has 50 sites (colored points in the XY space), with distinct local environments that select for diverse patterns of species diversity (size of points) and average trait values (color of points; results are shown at time $t=0, 10, 200$ for a simulation with $d=10^-3$ and $h^2 = 0.1$). (b) Regional metacommunity $\\Gamma$ diversity varied across levels of dispersal and heritability ($h^2$). (c) Variation partition plots after fitting metacommunity time series of species abundances $ln(N_{ikt})$ to HMSC models with environment, trait evolution, and random effects (space and time) as predictors. The top panel is for $h^2 = 0.1$, the middle panel is for $h^2 = 0.3$, and the bottom panel is for $h^2 = 0.9$."), echo=FALSE}
knitr::include_graphics("../output/fig2_mod.pdf",error=FALSE)
```

@@

The results for the HMSC analysis of the empirical data (community dynamics and frond size of four aquatic plant species under different environmental conditions) estimated an effect of trait change for species density. We used the HMSC models to calculate the mean proportion of variance explained (variance partitioning, VP) over all species, to estimate the relative importance of evolutionary (change in the frond size trait) and ecological (light and nutrient treatments) effects on species densities.Almost one fifth (18.6%) of the variance in species density could be explained by the variance in $|\Delta x|$ (\@ref(fig:fig3)a). Including trait evolution in the statistical model accurately led to reductions for the estimated importance of ecological effects by a total of 14.4% (from 37% to 26.6% for the amount of shading and from 31% to 27.1% for the nutrient concentrations) and the reduction of 4.2% of the random effects (from 18.7% to 15% for time and from 13.3% to 12.8% for replicates) (\@ref(fig:fig3)). While $|\Delta x|$ of the species Lm, Sp and Wc showed a similarly low impact (3.7%, 3.4%, and 3.7% respectively), $|\Delta x|$ of species Rn showed a higher impact on species density, 7.8%, which can be explained by its high $|\Delta x|$ changes over time (see Appendix S4 data). The 95% HPDI for main effect sizes showed a negative effect of light shading for species density, while nutrient concentration had a positive estimated effect for density (\@ref(fig:fig3)c). Changes in $|\Delta x|$ impacted densities in distinct ways for intra- and interspecific comparisons. The frond size of species Rn had a negative impact on Rn density, while change in Wc frond size had a positive estimated effect on species Sp and Lm (\@ref(fig:fig3)c). We show the estimated effect of trait evolution for species counts in \@ref(fig:fig3)d, which reflects potentially strong impacts in particular species, density, and light treatments and weak effects with high uncertainty for many cases.

```{r fig3, out.width="110%", include=TRUE, fig.align="center", fig.cap=c("The estimated effects of species trait changes and environmental treatments for densities of experimental aquatic plants (species Lm, Rn, Sp, and Wc). The variation partition for the HMSC model with (a) and without (b) trait change $|\\Delta x|$ is shown, as well as (c) the posterior estimated direction of effect (negative or positive; effect sizes with 95% HPDI that included 0 are colored white). The results indicated negative impacts of light, positive impacts of nutrients, negative intraspecific effects of $|\\Delta x_{Rn}|$ for $log(N_{Rn})$, negative interspecific effects of $|\\Delta x_{Lm}|$ for $log(N_{Sp})$, and positive interspecific effects of $|\\Delta x_{Wc}|$ for $log(N_{Sp})$ and $log(N_{Lm})$. (d) The posterior effect size estimates were used to estimate model-predicted densities at each time step of the experiment as a function only of $|\\Delta x|$ (for each of the four species, averaged across all of the other treatment effects). This captures the expected change in density due to trait evolution in all species."), echo=FALSE}
knitr::include_graphics("../output/fig3.pdf",error=FALSE)
```

@@

The analysis of community composition often requires statistical models that can consider the combined influences of environmental drivers and spatial connectivity. A more difficult, yet critical, element to incorporate is species interactions. Although the causal role of species interactions is difficult to interpret from statistical models (but see [@Dubart2022] [@Luo2022]) in systems with few species), joint species distribution models (JSDMs) have implemented spatial and temporal random effects that can effectively capture these species interactions using a latent variable approach (but note the resulting species covariance matrices will also reflect other processes besides species interactions; [@Poggiato2021]). Finally, one element that has not been fully implemented into joint species distribution models is dynamic trait evolution (see [@Garzon2019] for a model that does consider reaction norms from in-situ reciprocal transplant experiments). We used an existing JSDM, and developed simulation models of evolving populations in increasing complex scenarios to determine whether the signal of trait evolution could be estimated from data and compared to other drivers of community structure. We also applied this statistical model (HMSC; [@Ovaskainen2017HMSC] [@Tikhonov2020]) to an existing empirical dataset (from the experimental study of community eco-evolutionary dynamics in aquatic plants of [@Jewell2023]) to show how HMSC can be used to model the changes in species composition over time as a function of environmental drivers, trait change, and temporal random effects. Overall, there is no barrier to including dynamic trait change as a fixed effect in a JSDM model, and the resulting effect size coefficient estimates can be used to (i) estimate the relative importance of trait evolution for community composition and (ii) predict the impacts of future trait change in the system.

At the same time, fixed effects for the impacts of trait evolution on population size should be interpreted with a great deal of caution. For example, our simulations reveal instances where trait evolution may occur at the same time as large changes in population size, and this co-occurrence isn't always causal (e.g. when selection is weak but trait heritability is high, and species are introduced at low initial population size). In this instance, the regression coefficient from the model may be negative and strong. By using simulations of trait evolution and population growth in simple systems (1- and 2-species systems, at single and simplified multi-patch scales; Figure \@ref(fig:fig1); Appendix S1), we can better interpret effect size estimates. However, these precautions are not unique to including trait evolution as a predictor of community composition. For example, previous research has shown that the effect size coefficients from linear and AR models using $N_{i,t-1}$ to predict $N_{i,t}$ or $N_{j,t}$ are not equivalent to the intra- and interspecific interaction coefficients that determine species density dependence and should not automatically be used to interpret the direction and magnitude of species interactions ([@Kloppers2013] [@Certain2018] [@Muhlbauer2020] [@Olivenca2021]). Our findings for using HMSC to estimate the impact of trait evolution ($|\Delta x_i|$) for species abundances are similar - (i) the relative importance of trait evolution estimated by a variation partition accurately reflects the relative strength of simulated species interactions vs. magnitude of trait change and (ii) the coefficients greatly increase the predictive ability of the statistical model for simulated data (Appendix S1). HMSC is therefore a useful tool for understanding the role that evolution can play in community dynamics.

This was also observed when using HMSC to evaluate the importance of evolution for metacommunity structure, using simulated data from a model of population growth, competition, dispersal, and trait evolution in a spatially structured habitat (Figure \@ref(fig:fig1) a). First of all, the simulations themselves revealed an intriguing role of adaptive evolution for rescuing individual species from extinction (i.e. [@Gomulkiewicz1995]) and as a consequence greatly increasing biodiversity at the local and the landscape scale (i.e. community rescue, [@Low2015] [@Bell2019]; Figure \@ref(fig:fig2) b). The recovery of species via local adaptation had consequent effects for coexistence, as species that were initially mal-adapted could recover in population size and exert strong competitive effects on other species. Because HMSC can accurately estimate the relative importance of trait evolution for community composition (Appendix S3), we were thus able to test hypotheses about the drivers of increased or decreased impacts of trait evolution for community structure. Our simulation models fit to HMSC confirmed our hypotheses that the importance of evolution would decrease with increasing landscape connectivity (modeled as species dispersal rate $d$) and the overall speed of adaptive evolution (modeled as heritability $h^2$). Previous theoretical models have been shown that these two properties can have a strong influence on resulting community composition ([@Urban2008] [@Loeuille2008] [@Urban2009] [@DeMeester2016]), but our demonstration that the HMSC statistical model can accurately estimate $R^2$ values for the relative importance of trait evolution can allow hypotheses from theoretical models to be tested using empirical data and compared across different study systems.

The HMSC model was successfully applied to experimental aquatic plants, to estimate the importance of trait change (frond size), light levels, and nutrient composition for abundances of four species (Lm, Rn, Sp and Wc) over 11 weeks. This model currently does not take full advantage of the reciprocal transplant data (as implemented in the method of @Jewell2023) and thus doesn't separate the plastic elements of trait change (we also use species densities as our response, not the community average trait value). However, the HMSC model doesn't require derivation of new categories for effect sizes (this is the case for @Pantel2015, @Hattich2022, and @Jewell2023), nor must data be modelled as normal. As we show in Appendix S2 (also in @Ovaskainen2020), changes in population growth rate can also be fit as response variables, which may be more useful for mechanistic inference in some analyses. We also show the additional advantage of HMSC, which is the ability to predict species responses to environmental gradients and trait change. It should therefore be possible to use models such as HMSC to study how trait change impacts species densities under changing environmental conditions, or how the impact of trait change might depend on specific environmental conditions. These interactions are possible to consider in the Bayesian HMSC model.

As observed in the simulation results, correctly interpreting the estimated posterior effect sizes of evolution requires a thorough understanding of the experimental system. For example we often observed a negative effect of trait change for species density in model simulations, and this was also observed in the effect of trait change for intraspecific abundances in species. However these results derived from different mechanisms. In the simulations, trait change was greater when a species trait was further away from the optimal trait value ($E_k - x_{ikt}$), where they also had lower growth rates (the measure of fitness in our models) and thus lower population size. Species with lower trait distance ($E_k - x_{ikt}$) and thus higher population size had smaller trait changes. For this reason, HMSC estimates a negative effect of trait change $|\Delta x|$ for population size ($N_{ikt}$). Such effects should also be observed in studies of evolutionary rescue, where a species adapts to an initially lethal environment and avoids extinction via a gradually increasing growth rate ([@Gomulkiewicz1995]; demonstrated in Appendix S3). In the empirical data this negative effect size estimate resulted as the species Rn invested early in the experiment in a larger frond size (see Appendix S4). This may have been a tradeoff with a lower investment into reproduction, while still maintaining a similar total frond area on the water surface to capture the reduced light.

There has been a strong general interest in estimating effect sizes of evolution for ecological processes [@Hairston2005]; [@Ellner2011]; [@Pantel2015]; [@Barbour2022]), but in many empirical studies these estimates focus on one or a few species where the impacts of their evolution is strong enough to play a clear and detectable role in the corresponding ecological response variable of interest (e.g. population growth rate, community diversity, likelihood of food web collapse). There is no general knowledge for overall distributions of effect sizes for the impact of evolution for species composition in communities, nor are there studies that evaluate drivers of shifts in these distributions of effect sizes in complex communities (see [@Hermann2024] for estimates in a predator-prey system). We present here distributions of effect sizes of evolution for species abundances (i.e. Figure \@ref(fig:fig1) c,f,i; Figure \@ref(fig:fig2) c; Figure \@ref(fig:fig3) a,b) in simulated and experimental aquatic plant communities, and expect that these effect size coefficients can be useful for better understanding the relative importance of evolution as a driver of community composition.
